Practical Optional Types for Clojure

### 2016-01-14 intro.tex
colors

\begin{figure}[t!]
\inputminted[firstline=15]{clojure}{code/demo/src/demo/ex-intro.clj}
\caption{A simple Typed Clojure program}
\label{fig:ex1}
\end{figure}

>>>>

\begin{figure*}[t!]
  \normalsize
\begin{lstlisting}
(*typed ann pname [(U File String) -> (U nil String)] typed*)
(defmulti pname class)  ; multimethod dispatching on class of argument
(defmethod pname String [s] (*invoke pname (*interop new File s interop*) invoke*)) ; String case 
(defmethod pname File [f] (*interop .getName f interop*)) ; File case, static null check
(*invoke pname "STAINS/JELLY" invoke*) ;=> "JELLY" :- (U nil Str)
\end{lstlisting}
\caption{A simple Typed Clojure program (delimiter colors: {\color{interop}Java interoperation}, {\color{types}type annotation},
  {\color{invoke}function invocation}, {\color{red}collection literals}, {\color{mygray}miscellaneous})}
\label{fig:ex1}
\end{figure*}

### 2016-01-14 overview.tex
colors

\begin{lstlisting}
(ann  greet [Str -> Str])
(defn greet [n] (str "Hello, " n "!"))
(greet "Grace") ;=> "Hello, Grace!" :- Str
\end{lstlisting}
%
Providing \clj{nil} (exactly Java's \java{null})
is a static type error---\clj{nil} is not a string.
%
\begin{lstlisting}
(greet nil) ; Type Error: Expected Str, given nil
\end{lstlisting}

\paragraph{Unions} To allow \clj{nil}, we use \emph{ad-hoc unions} (\clj{nil} and \clj{false}
are logically false).
%
\begin{lstlisting}
(ann  greet-nil [(U nil Str) -> Str])
(defn greet-nil [n] (str "Hello" (when n (str ", " n)) "!"))
(greet-nil "Donald") ;=> "Hello, Donald!" :- Str 
(greet-nil nil)      ;=> "Hello!"         :- Str
\end{lstlisting}
%
%
Typed Clojure prevents well-typed code from dereferencing \clj{nil}.
%This is important for Clojure programs---\clj{nil}
%is treated like any other distinct datum in Clojure.

\paragraph{Flow analysis} Occurrence typing~\cite{TF10}
models type-based control flow.
In \clj{greetings}, a branch ensures \clj{repeat}
is never passed \clj{nil}.
%
\begin{lstlisting}
(ann  greetings [Str (U nil Int) -> Str])
(defn greetings [n i]
  (str "Hello, " (when i (apply str (repeat i "hello, "))) n "!"))
(greetings "Donald" 2)  ;=> "Hello, hello, hello, Donald!" :- Str
(greetings "Grace" nil) ;=> "Hello, Grace!"                :- Str
\end{lstlisting}
%
Removing the branch is a static type error---\clj{repeat} cannot be passed \clj{nil}.
%
\begin{lstlisting}
(ann  greetings-bad [Str (U nil Int) -> Str])
(defn greetings-bad [n i]           ; Expected Int, given (U nil Int)
  (str "Hello, " (apply str (repeat i "hello, ")) n "!"))
\end{lstlisting}

@@

\begin{exmp}
\begin{lstlisting}
(.getParent (new File "a/b"))  ;=> "a" :- (U nil Str)
\end{lstlisting}
\label{example:getparent-direct-constructor}
\end{exmp}
%
Typed Clojure can integrate with the Clojure compiler to avoid expensive reflective 
calls like \clj{getParent}, however if a specific overload cannot be found based on the
surrounding static context, a type error is thrown.
%
\begin{lstlisting}
(fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent
\end{lstlisting}
%
Function arguments default to \clj{Any}, which is similar to a union of all types. Ascribing
a parameter type allows Typed Clojure to find a specific method.

@@

\begin{exmp}
\begin{lstlisting}
(ann parent [(U nil File) -> (U nil Str)])
(defn parent [f] (if f (.getParent f) nil))
\end{lstlisting}
\label{example:parent-if}
\end{exmp}

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{lstlisting}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)) ; Type Error: Cannot call instance method on nil.
\end{lstlisting}

Typed Clojure rejects programs that assume methods cannot return \clj{nil}.
%
\begin{lstlisting}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f)) ; Type Error: Expected Str, given (U nil Str).
\end{lstlisting}
Method targets can never be \clj{nil}.
Typed Clojure also prevents passing \clj{nil} as Java method or
constructor arguments by default---this restriction can be
adjusted per method.

%
%Typed Clojure and Java treat \java{null} differently.
%In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
%called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
%This means the Java static type \java{String} is equivalent to
%\clj{(U nil String)} in Typed Clojure.
%
%Reference types in Java are nullable, so to guarantee a method call does not
%leak \java{null} into a Typed Clojure program we
%must assume methods can return \clj{nil}.
%
In contrast, JVM invariants guarantee constructors return non-null.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}
%
\begin{exmp}
\begin{lstlisting}
(parent (new File s))
\end{lstlisting}
\end{exmp}

@@

\begin{exmp}
\begin{lstlisting}
(ann hi [Kw -> Str]) ; multimethod type
(defmulti hi identity) ; dispatch function `identity`
(defmethod hi :en [_] "hello") ; method for `:en`
(defmethod hi :fr [_] "bonjour") ; method for `:fr`
(defmethod hi :default [_] "um...") ; default method
\end{lstlisting}
\label{example:hi-multimethod}
\end{exmp}

When invoked, the arguments are first supplied to the dispatch function---\clj{identity}---yielding
a \emph{dispatch value}. A method is then chosen
based on the dispatch value, to which the arguments are then passed to return a value.
%
\begin{lstlisting}
(map hi [:en :fr :bocce]) ;=> ("hello" "bonjour" "um...")
\end{lstlisting}
%
For example, 
\clj{(hi :en)} evaluates to \clj{"hello"}---it executes
the \clj{:en} method
because \clj{(= (identity :en) :en)} is true
and \clj{(= (identity :en) :fr)} is false.

@@

%(isa? (class "STAINS/JELLY") Object) ;=> true
%(isa? (identity :en) :fr) ;=> false
%(isa? (class (new File "JELLY")) String) ;=> false
\begin{lstlisting}
(isa? :en :en)       ;=> true
(isa? String Object) ;=> true
\end{lstlisting}
%
The current dispatch value and---in turn---each method's associated dispatch value
is supplied to \clj{isa?}. If exactly one method returns true, it is chosen.
%
For example,
the call
  \clj{(pname "STAINS/JELLY")}
picks the \clj{String} method because \clj{(isa? String String)}
is true, and
\clj{(isa? String File)}
is not.
%---\clj{(class "STAINS/JELLY")}
%is \clj{String}.
%
%The \clj{String} method body
%\clj{(pname (new File "STAINS/JELLY"))}
%chooses the \clj{File} method for opposite reasons.

@@

%Beyond primitives and Java objects, 
The most common way to represent compound data in Clojure 
are immutable hash-maps, typicially with keyword keys.
%
Keywords double as functions that
look themselves up in a map, or return \clj{nil} if absent.
%
\begin{exmp}
\begin{lstlisting}
(def breakfast {:en "waffles" :fr "croissants"})
(:en breakfast)    ;=> "waffles" :- Str
(:bocce breakfast) ;=> nil       :- nil
\end{lstlisting}
\label{example:breakfastcomplete}
\end{exmp}

@@

HMap types default to partial specification, with
\clj{\{:en Str :fr Str\}} abbreviating \clj{(HMap :mandatory \{:en Str, :fr Str\})}.
%
\begin{exmp}
\begin{lstlisting}
(ann lunch '{:en Str :fr Str})
(def lunch {:en "muffin" :fr "baguette"})
(:bocce lunch) ;=> nil :- Any ; less accurate type
\end{lstlisting}
\label{example:lunchpartial}
\end{exmp}
%(:en lunch)    ; :- Str
%;=> "muffin"
%(:fr lunch)    ; :- Str
%;=> "baguette"

@@

\begin{exmp}
\begin{lstlisting}
(defalias RawKeyPair ; extra keys disallowed
  (HMap :mandatory {:pub RawKey, :priv RawKey}, 
        :complete? true))
(defalias EncKeyPair ; extra keys disallowed
  (HMap :mandatory {:pub RawKey, :enc-priv EncKey}, :complete? true))

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair [kp]
  (assoc (dissoc kp :priv) :enc-priv (encrypt (:priv kp))))
\end{lstlisting}
\label{example:circleci}
\end{exmp}
%
%(ann enc-keypair [RawKeyPair -> EncKeyPair])
%(defn enc-keypair [{pk :priv :as kp}] ; original map is kp
%  (assoc (dissoc kp :priv)       ; remove unencrypted private key
%         :enc-priv (encrypt pk))) ; add encrypted private key
%
%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
As \clj{EncKeyPair} is fully specified, we remove extra keys like \clj{:priv}
via \clj{dissoc}, which returns a new map that is the first argument without the
entry named by the second argument. Notice removing \clj{dissoc} causes a type error.
%
\begin{lstlisting}
(defn enc-keypair-bad [kp] ; Type error: :priv disallowed
  (assoc kp :enc-priv (encrypt (:priv kp))))
\end{lstlisting}

@@

\begin{lstlisting}
(defalias Order "A meal order, tracking dessert quantities."
  (U '{:Meal ':lunch, :desserts Int} '{:Meal ':dinner :desserts Int}
     '{:Meal ':combo :meal1 Order :meal2 Order}))
\end{lstlisting}

The \clj{:Meal} entry is common to each HMap, always mapped to a known keyword singleton
type.
It's natural to dispatch on the \clj{class} of an instance---it's similarly
natural to dispatch on a known entry like \clj{:Meal}.
%
\begin{exmp2} % use other environment to work around page break
\begin{lstlisting}
(ann desserts [Order -> Int])
(defmulti desserts :Meal)  ; dispatch on :Meal entry
(defmethod desserts :lunch [o] (:desserts d))
(defmethod desserts :dinner [o] (:desserts d))
(defmethod desserts :combo [o] 
  (+ (desserts (:meal1 o)) (desserts (:meal2 o))))

(desserts {:Meal :combo, :meal1 {:Meal :lunch :desserts 1}, 
           :meal2 {:Meal :dinner :desserts 2}}) ;=> 3
\end{lstlisting}
\label{example:desserts-on-meal}
\end{exmp2}
%
The \clj{:combo} method is verified to only structurally recur
on \clj{Order}s. This is achieved because we learn the argument \clj{o} must
be of type
\clj{'{:Meal :combo}}
since
\clj{(isa? (:Meal o) :combo)}
is true. Combining this
with the fact that \clj{o} is an \clj{Order}
eliminates possibility of \clj{:lunch} and \clj{:dinner}
orders, simplifying \clj{o} to
\clj{'{:Meal ':combo :meal1 Order :meal2 Order}}
which contains appropriate arguments for both recursive calls.

@@

\begin{exmp}
\begin{lstlisting}
(ann desserts' [Order -> Int])
(defmulti desserts' 
  (fn [o :- Order] (class (:desserts o))))
(defmethod desserts' Long [o] 
; o :- (U '{:Meal (U ':dinner ':lunch), :desserts Int}
;       '{:Meal ':combo, :desserts Int, :meal1 Order, :meal2 Order})
  (:desserts o))
(defmethod desserts' nil [o]
  ; o :- '{:Meal ':combo, :meal1 Order, :meal2 Order}
  (+ (desserts' (:meal1 o)) (desserts' (:meal2 o))))
\end{lstlisting}
\label{example:desserts-on-class}
\end{exmp}
%
%(desserts' {:Meal :combo 
%            :meal1 {:Meal :lunch :desserts 1}
%            :meal2 {:Meal :dinner :desserts 2}})
%;=> 3

In the \clj{Long} method, Typed Clojure learns that
its argument is at least of type \clj{'{:desserts Long}}---since
\clj{(isa? (class (:desserts o)) Long)}
must be true.
%
%Knowing \clj{o} is also an
%\clj{Order},
Here the
\clj{:desserts} entry
\emph{must} be present and mapped to a \clj{Long}---even in a \clj{:combo} meal,
which does not specify \clj{:desserts}
as present or absent.

In the \clj{nil} method,
\clj{(isa? (class (:desserts o)) nil)}
must be true---which implies \clj{(class (:desserts o))} is \clj{nil}.
%
Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, like \clj{{:desserts nil}}, or
  \item \clj{o} is missing a \clj{:desserts} entry.%, like \clj{{}}.
\end{itemize}
We can express this type with the \clj{:absent-keys} HMap option
%Equivalently, \clj{o} is of type
% Note: mintedinline doesn't work with hash characters #
\begin{lstlisting}
(U '{:desserts nil} (HMap :absent-keys #{:desserts}))
\end{lstlisting}
This eliminates non-\clj{:combo} meals
since their \clj{'{:desserts Int}} type does not agree
with this new information (because \clj{:desserts}
is neither \clj{nil} or absent).

>>>>

\begin{lstlisting}
(*typed ann  greet [Str -> Str] typed*)
(defn greet [n] (*invoke str "Hello, " n "!" invoke*))
(*invoke greet "Grace" invoke*) ;=> "Hello, Grace!" :- Str
\end{lstlisting}
%
Providing \clj{nil} (exactly Java's \java{null})
is a static type error---\clj{nil} is not a string.
%
\begin{lstlisting}
(*invoke greet nil invoke*) ; Type Error: Expected Str, given nil
\end{lstlisting}

\paragraph{Unions} To allow \clj{nil}, we use \emph{ad-hoc unions} (\clj{nil} and \clj{false}
are logically false).
%
\begin{lstlisting}
(*typed ann  greet-nil [(U nil Str) -> Str] typed*)
(defn greet-nil [n] (*invoke str "Hello" (when n (*invoke str ", " n invoke*)) "!" invoke*))
(*invoke greet-nil "Donald" invoke*) ;=> "Hello, Donald!" :- Str 
(*invoke greet-nil nil invoke*)      ;=> "Hello!"         :- Str
\end{lstlisting}
%
%
Typed Clojure prevents well-typed code from dereferencing \clj{nil}.
%This is important for Clojure programs---\clj{nil}
%is treated like any other distinct datum in Clojure.

\paragraph{Flow analysis} Occurrence typing~\cite{TF10}
models type-based control flow.
In \clj{greetings}, a branch ensures \clj{repeat}
is never passed \clj{nil}.
%
\begin{lstlisting}
(*typed ann  greetings [Str (U nil Int) -> Str] typed*)
(defn greetings [n i]
  (*invoke str "Hello, " (when i (*invoke apply str (*invoke repeat i "hello, " invoke*) invoke*)) n "!" invoke*))
(*invoke greetings "Donald" 2 invoke*)  ;=> "Hello, hello, hello, Donald!" :- Str
(*invoke greetings "Grace" nil invoke*) ;=> "Hello, Grace!"                :- Str
\end{lstlisting}
%
Removing the branch is a static type error---\clj{repeat} cannot be passed \clj{nil}.
%
\begin{lstlisting}
(*typed ann  greetings-bad [Str (U nil Int) -> Str] typed*)
(defn greetings-bad [n i]           ; Expected Int, given (U nil Int)
  (*invoke str "Hello, " (*invoke apply str (*invoke repeat i "hello, " invoke*) invoke*) n "!" invoke*))
\end{lstlisting}

@@

\begin{exmp}
\begin{lstlisting}
(*interop .getParent (*interop new File "a/b" interop*) interop*)  ;=> "a" :- (U nil Str)
\end{lstlisting}
\label{example:getparent-direct-constructor}
\end{exmp}
%
Typed Clojure can integrate with the Clojure compiler to avoid expensive reflective 
calls like \clj{getParent}, however if a specific overload cannot be found based on the
surrounding static context, a type error is thrown.
%
\begin{lstlisting}
(fn [f] (*interop .getParent f interop*)) ; Type Error: Unresolved interop: getParent
\end{lstlisting}
%
Function arguments default to \clj{Any}, which is similar to a union of all types. Ascribing
a parameter type allows Typed Clojure to find a specific method.

@@

\begin{exmp}
\begin{lstlisting}
(*typed ann parent [(U nil File) -> (U nil Str)] typed*)
(defn parent [f] (if f (*interop .getParent f interop*) nil))
\end{lstlisting}
\label{example:parent-if}
\end{exmp}

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{lstlisting}
(defn parent-bad-in [f :- (U nil File)]
  (*interop .getParent f interop*)) ; Type Error: Cannot call instance method on nil.
\end{lstlisting}

Typed Clojure rejects programs that assume methods cannot return \clj{nil}.
%
\begin{lstlisting}
(defn parent-bad-out [f :- File] :- Str
  (*interop .getParent f interop*)) ; Type Error: Expected Str, given (U nil Str).
\end{lstlisting}
Method targets can never be \clj{nil}.
Typed Clojure also prevents passing \clj{nil} as Java method or
constructor arguments by default---this restriction can be
adjusted per method.

%
%Typed Clojure and Java treat \java{null} differently.
%In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
%called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
%This means the Java static type \java{String} is equivalent to
%\clj{(U nil String)} in Typed Clojure.
%
%Reference types in Java are nullable, so to guarantee a method call does not
%leak \java{null} into a Typed Clojure program we
%must assume methods can return \clj{nil}.
%
In contrast, JVM invariants guarantee constructors return non-null.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}
%
\begin{exmp}
\begin{lstlisting}
(*invoke parent (*interop new File s interop*) invoke*)
\end{lstlisting}
\end{exmp}

@@

\begin{exmp}
\begin{lstlisting}
(*typed ann hi [Kw -> Str] typed*) ; multimethod type
(defmulti hi identity) ; dispatch function `identity`
(defmethod hi :en [_] "hello") ; method for `:en`
(defmethod hi :fr [_] "bonjour") ; method for `:fr`
(defmethod hi :default [_] "um...") ; default method
\end{lstlisting}
\label{example:hi-multimethod}
\end{exmp}

When invoked, the arguments are first supplied to the dispatch function---\clj{identity}---yielding
a \emph{dispatch value}. A method is then chosen
based on the dispatch value, to which the arguments are then passed to return a value.
%
\begin{lstlisting}
(*invoke map hi [*vec :en :fr :bocce vec*] invoke*) ;=> (*list "hello" "bonjour" "um..." list*)
\end{lstlisting}
%
For example, 
\clj{(*invoke hi :en invoke*)} evaluates to \clj{"hello"}---it executes
the \clj{:en} method
because \clj{(*invoke = (*invoke identity :en invoke*) :en invoke*)} is true
and \clj{(*invoke = (*invoke identity :en invoke*) :fr invoke*)} is false.

@@

%(isa? (class "STAINS/JELLY") Object) ;=> true
%(isa? (identity :en) :fr) ;=> false
%(isa? (class (new File "JELLY")) String) ;=> false
\begin{lstlisting}
(*invoke isa? :en :en invoke*)       ;=> true
(*invoke isa? String Object invoke*) ;=> true
\end{lstlisting}
%
The current dispatch value and---in turn---each method's associated dispatch value
is supplied to \clj{isa?}. If exactly one method returns true, it is chosen.
%
For example,
the call
  \clj{(*invoke pname "STAINS/JELLY" invoke*)}
picks the \clj{String} method because \clj{(*invoke isa? String String invoke*)}
is true, and
\clj{(*invoke isa? String File invoke*)}
is not.
%---\clj{(class "STAINS/JELLY")}
%is \clj{String}.
%
%The \clj{String} method body
%\clj{(pname (new File "STAINS/JELLY"))}
%chooses the \clj{File} method for opposite reasons.

@@

%Beyond primitives and Java objects, 
The most common way to represent compound data in Clojure 
are immutable hash-maps, typicially with keyword keys.
%
Keywords double as functions that
look themselves up in a map, or return \clj{nil} if absent.
%
\begin{exmp}
\begin{lstlisting}
(def breakfast {*map :en "waffles" :fr "croissants" map*})
(*invoke :en breakfast invoke*)    ;=> "waffles" :- Str
(*invoke :bocce breakfast invoke*) ;=> nil       :- nil
\end{lstlisting}
\label{example:breakfastcomplete}
\end{exmp}

@@

HMap types default to partial specification, with
\clj{'\{:en Str :fr Str\}} abbreviating \clj{(HMap :mandatory \{:en Str, :fr Str\})}.
%
\begin{exmp}
\begin{lstlisting}
(*typed ann lunch '{:en Str :fr Str} typed*)
(def lunch {*map :en "muffin" :fr "baguette" map*})
(*invoke :bocce lunch invoke*) ;=> nil :- Any ; less accurate type
\end{lstlisting}
\label{example:lunchpartial}
\end{exmp}
%(:en lunch)    ; :- Str
%;=> "muffin"
%(:fr lunch)    ; :- Str
%;=> "baguette"

@@

\begin{exmp}
\begin{lstlisting}
(*typed defalias RawKeyPair ; extra keys disallowed
  (HMap :mandatory {:pub RawKey, :priv RawKey}, 
        :complete? true) typed*)
(*typed defalias EncKeyPair ; extra keys disallowed
  (HMap :mandatory {:pub RawKey, :enc-priv EncKey}, :complete? true) typed*)

(*typed ann enc-keypair [RawKeyPair -> EncKeyPair] typed*)
(defn enc-keypair [kp]
  (*invoke assoc (*invoke dissoc kp :priv invoke*) :enc-priv (*invoke encrypt (*invoke :priv kp invoke*) invoke*) invoke*))
\end{lstlisting}
\label{example:circleci}
\end{exmp}
%
%(ann enc-keypair [RawKeyPair -> EncKeyPair])
%(defn enc-keypair [{pk :priv :as kp}] ; original map is kp
%  (assoc (dissoc kp :priv)       ; remove unencrypted private key
%         :enc-priv (encrypt pk))) ; add encrypted private key
%
%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
As \clj{EncKeyPair} is fully specified, we remove extra keys like \clj{:priv}
via \clj{dissoc}, which returns a new map that is the first argument without the
entry named by the second argument. Notice removing \clj{dissoc} causes a type error.
%
\begin{lstlisting}
(defn enc-keypair-bad [kp] ; Type error: :priv disallowed
  (*invoke assoc kp :enc-priv (*invoke encrypt (*invoke :priv kp invoke*) invoke*) invoke*))
\end{lstlisting}

@@

\begin{lstlisting}
(*typed defalias Order "A meal order, tracking dessert quantities."
  (U '{:Meal ':lunch, :desserts Int} '{:Meal ':dinner :desserts Int}
     '{:Meal ':combo :meal1 Order :meal2 Order}) typed*)
\end{lstlisting}

The \clj{:Meal} entry is common to each HMap, always mapped to a known keyword singleton
type.
It's natural to dispatch on the \clj{class} of an instance---it's similarly
natural to dispatch on a known entry like \clj{:Meal}.
%
\begin{exmp2} % use other environment to work around page break
\begin{lstlisting}
(*typed ann desserts [Order -> Int] typed*)
(defmulti desserts :Meal)  ; dispatch on :Meal entry
(defmethod desserts :lunch [o] (*invoke :desserts d invoke*))
(defmethod desserts :dinner [o] (*invoke :desserts d invoke*))
(defmethod desserts :combo [o] 
  (*invoke + (*invoke desserts (*invoke :meal1 o invoke*) invoke*) (*invoke desserts (*invoke :meal2 o invoke*) invoke*) invoke*))

(*invoke desserts {*map :Meal :combo, :meal1 {*map :Meal :lunch :desserts 1 map*}, 
           :meal2 {*map :Meal :dinner :desserts 2 map*} map*} invoke*) ;=> 3
\end{lstlisting}
\label{example:desserts-on-meal}
\end{exmp2}
%
The \clj{:combo} method is verified to only structurally recur
on \clj{Order}s. This is achieved because we learn the argument \clj{o} must
be of type
\clj{'\{:Meal :combo\}}
since
\clj{(isa? (:Meal o) :combo)}
is true. Combining this
with the fact that \clj{o} is an \clj{Order}
eliminates possibility of \clj{:lunch} and \clj{:dinner}
orders, simplifying \clj{o} to
\clj{'\{:Meal ':combo :meal1 Order :meal2 Order\}}
which contains appropriate arguments for both recursive calls.

@@

\begin{exmp}
\begin{lstlisting}
(*typed ann desserts' [Order -> Int] typed*)
(defmulti desserts' 
  (fn [o :- Order] (*invoke class (*invoke :desserts o invoke*) invoke*)))
(defmethod desserts' Long [o] 
;o :- (U '{:Meal (U ':dinner ':lunch), :desserts Int}
;       '{:Meal ':combo, :desserts Int, :meal1 Order, :meal2 Order})
  (*invoke :desserts o invoke*))
(defmethod desserts' nil [o]
  ; o :- '{:Meal ':combo, :meal1 Order, :meal2 Order}
  (*invoke + (*invoke desserts' (*invoke :meal1 o invoke*) invoke*) (*invoke desserts' (*invoke :meal2 o invoke*) invoke*) invoke*))
\end{lstlisting}
\label{example:desserts-on-class}
\end{exmp}
%
%(desserts' {:Meal :combo 
%            :meal1 {:Meal :lunch :desserts 1}
%            :meal2 {:Meal :dinner :desserts 2}})
%;=> 3

In the \clj{Long} method, Typed Clojure learns that
its argument is at least of type \clj{'\{:desserts Long\}}---since
\clj{(*invoke isa? (*invoke class (*invoke :desserts o invoke*) invoke*) Long invoke*)}
must be true.
%
%Knowing \clj{o} is also an
%\clj{Order},
Here the
\clj{:desserts} entry
\emph{must} be present and mapped to a \clj{Long}---even in a \clj{:combo} meal,
which does not specify \clj{:desserts}
as present or absent.

In the \clj{nil} method,
\clj{(*invoke isa? (*invoke class (*invoke :desserts o invoke*) invoke*) nil invoke*)}
must be true---which implies \clj{(*invoke class (*invoke :desserts o invoke*) invoke*)} is \clj{nil}.
%
Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, like \clj{{:desserts nil}}, or
  \item \clj{o} is missing a \clj{:desserts} entry.%, like \clj{{}}.
\end{itemize}
We can express this type with the \clj{:absent-keys} HMap option
%Equivalently, \clj{o} is of type
% Note: mintedinline doesn't work with hash characters #
\begin{lstlisting}
(U '{:desserts nil} (HMap :absent-keys #{:desserts}))
\end{lstlisting}
This eliminates non-\clj{:combo} meals
since their \clj{'\{:desserts Int\}} type does not agree
with this new information (because \clj{:desserts}
is neither \clj{nil} or absent).
